---
title: "CBM_defaults"
author: ""
date: "19 January 2018"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

# Overview

This module can run independently by running the global script below.
Its main task is to read-in all the default values in SpaDES-CBM which is akin to the `ArchiveIndex` in CBM-CFS3 runs.

# Usage

```{r module_usage}
library(igraph)
library(SpaDES.core)

cacheDir <- reproducible::checkPath("cache", create = TRUE)
moduleDir <- "modules"
inputDir <- reproducible::checkPath("inputs", create = TRUE)
outputDir <- reproducible::checkPath("outputs", create = TRUE)
setPaths(inputPath = inputDir, 
         modulePath = moduleDir, 
         outputPath = outputDir, 
         cachePath = cacheDir)

times <- list(start = 0, end = 10)
parameters <- list(
  #.progress = list(type = "text", interval = 1), # for a progress bar
  ## If there are further modules, each can have its own set of parameters:
  #module1 = list(param1 = value1, param2 = value2),
  #module2 = list(param1 = value1, param2 = value2)
)
modules <- list("CBM_defaults")
objects <- list()
paths <- list(
  cachePath = cacheDir,
  modulePath = moduleDir,
  inputPath = inputDir,
  outputPath = outputDir
)

myDefaults <- simInit(times = times, params = parameters, modules = modules,
                      objects = objects, paths = paths)

outDefaults <- spades(myDefaults)
```

```{r tests}
# The spades call should make a "dataset" class object
testthat::expect_true(is(outSim@.envir$cbmData, "dataset"))
```

# Events

Describe what happens for each event type.

## Plotting

Write what is plotted.

## Saving

Write what is saved.

# Data dependencies

## Input data

How to obtain input data, and a description of the data required by the module.
If `sourceURL` is specified, `downloadData("CBM_defaults", "path/to/modules/dir")` may be sufficient.

## Output data

Description of the module outputs.

# Links to other modules

Describe any anticipated linkages to other modules.

